When does vapor pressure deficit drive or reduce evapotranspiration?

### 2019-02-28 doc/paper/body.tex
done with pierre's comments, still need to put in soil moisture discussion and supplementary materials somewhere

Additionally, the sensitivity of ET response to plant parameters
highlights the importance of understanding and developing stomatal
conductance models and parameters for the ecosystem scale. In
particular, why and how g$_1$ behavior at the leaf scale is not
analagous to the ecosystem scale must be understood. While the derivation
here was explicit about the assumption of a constant g$_1$ term with
respect to the VPD perturbation, many sophisticated land surface
models and earth system models employing stomatal conductance models
make similar, and even stronger assumptions about the stationarity of
VPD slope terms. Many models will assume a constant g$_1$ term within
a given plant functional type, and research derived from these models
often does not acknowledge limitations of this assumption, or the
difficulty in the literature with quantifying an ecosystem scale slope
parameter \citep{Medlyn_2017}. Given that something so fundamental as
the ET response to VPD varies strongly with g$_1$, we must invest more
in understanding g$_1$'s behavior at the ecosystem scale, and the
efficacy of models using g$_1$'s for ecosystem-scale
fluxes. Fortunately, our framework, and the novel approach of
using robust ecosystem-scale ratios (uWUE) to remove the implicit
dependence of stomatal conductance on ET, is flexible enough to be
applied to any future-derived ecosystem-scale stomatal conductance
model that contains a dependence on GPP.

So far we have discussed the role of different parameters in
modulating the ET response. However, the form of the ET response is
imposed by our choice of stomatal conductance model. We now examine
how the model choice, and specifically the exponent of the VPD
dependence in the stomatal conductance model, can impact the general
form of the ET-VPD relationship. This is relevant as many models
include an exponentiated VPD term \citep[e.g.,][]{Leuning_1990}, but
the value of the exponent can vary. It is useful to understand how the
choice of model imposes the functional form of the ET-VPD
relationship, given its fundamental importance for land atmosphere
interaction.

@@

The results presented in Sect. \ref{results} indicates that for a
given $uWUE$ and $g_1$, the ET dependence on VPD should be concave
upward. In other words, there should be some local minimum in ET at a
critical VPD$_{crit}$, assuming the scaling and plant terms  (e.g. aerodynamic
conductance, $\Delta$, $g_1$ and $uWUE$) are held fixed. This result warrants further
investigation, because to our knowledge no one has derived the
theoretical ecosystem-scale relationship between ET and VPD while
controlling for other environmental conditions. In particular, from
personal communication, there is an apparent lack of consensus over
whether the shape of the ET-VPD curve should be concave upward (our
result) or concave downward in the absence of dramatic water stress. Given that understanding the ET-VPD
relationship of the one-way plant response is fundamental to
hypothesizing about any feedbacks between the land surface and the
atmosphere, we analyze why our derived ET-VPD relationship is concave
upward, particularly with respect to the exponent of VPD dependence in
$uWUE$ and the Medlyn unified stomatal conductance model.

There is a theoretical basis for the square root VPD dependence in
both the stomatal conductance model and $uWUE$ based on the assumption
that stomata behave to maximize carbon gain while minimizing water
loss, which observations also generally support \citep{Lloyd_1991,
  MEDLYN_2011, Lin_2015, Zhou_2014, Zhou_2015, Medlyn_2017}. However,
some purely empirical results that fit the exponent of the VPD
dependence to data have shown that it may vary slightly from 1/2,
suggesting that stomata, as well as ecosystem-scale quantities based
on stomata theory, may not always function optimally \citep{Zhou_2015,
  Lin_2018}. Specifically with regards to $uWUE$, one would not expect
that this ecosystem scale WUE quantity will respond to VPD exactly
analogously to stomata. Direct soil evaporation's contributions to ET
should shift the exponent of the VPD dependence, especially at
conditions of low GPP when we would expect a systematically larger
portion of direct soil evaporation contributions to ET, because we
would also expect lower amounts of transpiration at low GPP. \citet{Zhou_2015}'s
results corroborate this: they found a mean empirically fit
exponential VPD dependence of 0.55, varying slightly from the
theoretically optimal value of of 0.5 for AmeriFlux
sites. Interpreting \citet{Lin_2018}'s results, which also show
variance in the empirical exponent of the VPD dependence of the
stomatal conductance model, is more difficult as \citet{Lin_2018} do
not handle GPP/A dependence of stomatal conductance in a directly
analogous manner to the optimal theory in \citet{MEDLYN_2011} and
\citet{Medlyn_2017}. Regardless, given that these recent results on
the relationship between VPD, GPP, and ET \citep{MEDLYN_2011,
  Zhou_2014, Zhou_2015, Medlyn_2017} form the backbone of our analysis
and are what allowed us to derive an explicit ET expression for the
first time (Eq. (\ref{et})), we will analyze if and how assumptions
about the exponent of the VPD dependence impacts the shape of the
ET-VPD dependence. This analysis is also important to understand whether the choice of stomatal conductance model alters the fundamental behavior of the ET-VPD relationships, as many commonly used models  utilize a VPD exponent other than the 1/2 suggested by optimal theory \citep[e.g. ][ which uses an exponent of 1]{Leuning_1990}.

By introducing $n$ and $m$ we can free our stomatal
conductance model from assumptions about VPD dependence:
% \begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g*}{VPD^m}\right) \frac{*WUE \; ET}{c_a \; VPD^n},
    \label{m_n}
  \end{equation}
% \end{linenomath*}
where:
\[*WUE = \frac{GPP}{ET}VPD^n,\] and $g*$ is a generic slope parameter
of units $VPD^m$. To determine how the exponent $n$ and $m$ alter the
shape of the ET-VPD dependence we find the roots of the second
derivative of ET, using Eq. (\ref{m_n}) for stomatal conductance
($g_s$), with respect to VPD: \input{d2_solutions.tex} With this
result we have defined the family of curves separating concave up from
concave down ET solutions (Fig. \ref{concave}). These curves are
only functions of the exponent of the VPD dependence and a quantity we
call non-dimensional VPD ($VPD^m/g*$). Several important
relations reveal themselves from Eq. (\ref{curves}):

@@

While it is possible that in the future some other form of
VPD dependence is derived, at present \cite{MEDLYN_2011} and
\cite{Zhou_2014} firmly established n=m=1/2 as the most likely
candidate given current theory and empirical data. Additionally, we
argue that a concave up result matches physical intuition more than a
concave down result. Plants must maintain nutrient and sugar transport
through the phloem and xylem. To accomplish this, stomata must remain
slightly open \citep{De_2013, Nikinmaa_2013, Ryan_2014}. Furthermore,
even if complete stomatal closure were possible, cuticular water loss
and [at the ecosystem-scale] direct soil evaporation are still sources
of ET which increase with VPD, independent of stomatal
closure. Therefore, in the limit as VPD becomes large and we assume
plants are exercising all strategies to reduce ET, any further increase
in VPD should result in an increase in ET through cuticular water loss
and/or direct soil evaporation. This inevitable transition from
conditions when stomata respond strongly to VPD to conditions when
stomata response is asymptoting towards full closure would cause a
concave up ET-VPD curve, which is matched by the theory. In short,
plant response becomes more limited as VPD increases, while
atmospheric demand monotonically increases with VPD, leading to the
 result that atmospheric demand dominates plant response when
atmospheric demand is high.

This analysis allows us to understand the theoretical shape of the ET
response to VPD with environmental conditions held
fixed. Accomplishing this with purely statistical methods applied to
flux observations would be very difficult, given the relatively fast
time scale of plant response and the non-stationarity of [solar
forced] environmental conditions over the relatively coarse (half
hourly) flux estimates (which is required to obtain robust
eddy-covariance statistics). Our results on the shape of the ET-VPD
curve with environmental conditions held fixed can be built upon with
future work examining how changes in VPD and environmental conditions
(e.g. soil water storage) feedback upon one another. In the soil water
storage example, over very long time scales extremely high VPD
perturbations coupled with no precipitation could result in decreases
in soil water storage such that water becomes limiting. This could be
represented by an extension of our framework in which uWUE is allowed
to decrease with decreasing SWC. Here, we focus our results by
framing theme with previous literature to  build baseline intuition for
ET-VPD dependence.

>>>>

Additionally, the sensitivity of ET response to plant parameters
highlights the importance of understanding and developing stomatal
conductance models and parameters for the ecosystem scale. In
particular, why and how g$_1$ behavior at the leaf scale is not
analagous to the ecosystem scale must be understood. While the
derivation here was explicit about the assumption of a constant g$_1$
term with respect to the VPD perturbation, many sophisticated land
surface models and earth system models employing stomatal conductance
models make similar, and even stronger assumptions about the
stationarity of VPD slope terms \citep{Niu_2011, Franks_2017,
  Rogers_2017, Lawrence_2019}. Many models will assume a constant
g$_1$ term within a given plant functional type, and research derived
from these models often does not acknowledge limitations of this
assumption, or the difficulty in the literature with quantifying an
ecosystem scale slope parameter \citep{Medlyn_2017}. Given that
something so fundamental as the ET response to VPD varies strongly
with g$_1$, we must invest more in understanding g$_1$'s behavior at
the ecosystem scale, and the efficacy of models using g$_1$'s for
ecosystem-scale fluxes. Fortunately, our framework, and the novel
approach of using robust ecosystem-scale ratios (uWUE) to remove the
implicit dependence of stomatal conductance on ET, is flexible enough
to be applied to any stomatal conductance model that contains a
dependence on GPP.

So far we have discussed the role of different parameters in
modulating the ET response. However, the form of the ET response is
imposed by our choice of stomatal conductance model. We now examine
how the stomatal model choice, and specifically the exponent of the
VPD dependence in the stomatal conductance model, can impact the
general form of the ET-VPD relationship. There is a theoretical basis
for the square root VPD dependence in both the stomatal conductance
model and $uWUE$ based on the assumption that stomata behave to
maximize carbon gain while minimizing water loss, which observations
also generally support \citep{Lloyd_1991, MEDLYN_2011, Lin_2015,
  Zhou_2014, Zhou_2015, Medlyn_2017}. However, some purely empirical
results that fit the exponent of the VPD dependence to data have shown
that it may vary slightly from 1/2, suggesting that stomata, as well
as ecosystem-scale quantities based on stomata theory, may not always
function optimally \citep{Zhou_2015, Lin_2018}. Specifically with
regards to $uWUE$, one would not expect that this ecosystem scale WUE
quantity will respond to VPD exactly analogously to stomata. Direct
soil evaporation's contributions to ET should shift the exponent of
the VPD dependence, especially at conditions of low GPP when we would
expect a systematically larger portion of direct soil evaporation
contributions to ET, because we would also expect lower amounts of
transpiration at low GPP. \citet{Zhou_2015}'s results corroborate
this: they found a mean empirically fit exponential VPD dependence of
0.55, varying slightly from the theoretically optimal value of of 0.5
for AmeriFlux sites. \citet{Lin_2018}'s results also show variance in
the empirical exponent of the VPD dependence of the stomatal
conductance model, but interpretation of this variance is more
difficult as \citet{Lin_2018} do not handle GPP/A dependence of
stomatal conductance in a directly analogous manner to the optimal
theory in \citet{MEDLYN_2011} and \citet{Medlyn_2017}. Regardless,
given that these recent results on the relationship between VPD, GPP,
and ET \citep{MEDLYN_2011, Zhou_2014, Zhou_2015, Medlyn_2017} form the
backbone of our analysis and are what allowed us to derive an explicit
ET expression for the first time (Eq. (\ref{et})), we will analyze if
and how assumptions about the exponent of the VPD dependence impacts
the shape of the ET-VPD dependence. This analysis is also important to
understand whether the choice of stomatal conductance model alters the
fundamental behavior of the ET-VPD relationships, as many commonly
used models utilize a VPD exponent other than the 1/2 suggested by
optimal theory \citep[e.g. ][ which uses an exponent of
1]{Leuning_1990}. It is useful to understand how the choice of model
imposes the functional form of the ET-VPD relationship, given its
fundamental importance for land-atmosphere interactions.

@@

The results presented in Sect. \ref{results} indicates that for a
given $uWUE$ and $g_1$, the ET dependence on VPD should be concave
upward. In other words, there should be some local minimum in ET at a
critical VPD$_{crit}$, assuming the scaling and plant terms
(e.g. aerodynamic conductance, $\Delta$, $g_1$ and $uWUE$) are held
fixed. This result warrants further investigation, because to our
knowledge no earlier work has derived the theoretical ecosystem-scale
relationship between ET and VPD while controlling for other
environmental conditions. In particular there is an apparent lack of
consensus over whether the shape of the ET-VPD curve should be concave
upward (our result) or concave downward in the absence of dramatic
water stress. Given that understanding the ET-VPD relationship of the
one-way plant response is fundamental to hypothesizing about any
feedbacks between the land surface and the atmosphere, we analyze why
our derived ET-VPD relationship is concave upward, particularly with
respect to the exponent of VPD dependence in $uWUE$ and the Medlyn
unified stomatal conductance model, as other models and empiracle
results have suggested exponents varying from 1/2 \citep{Leuning_1990,
  Zhou_2015, Lin_2018}.

By introducing $n$, the exponent of VPD in uWUE, and $m$, the exponent
of VPD in the stomatal conductance model, we can free our theory from
assumptions about VPD dependence:
% \begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g*}{VPD^m}\right) \frac{*WUE \; ET}{c_a \; VPD^n},
    \label{m_n}
  \end{equation}
% \end{linenomath*}
where:
\[*WUE = \frac{GPP}{ET}VPD^n,\] and $g*$ is a generic slope parameter
of units $VPD^m$. To determine how the exponent $n$ and $m$ alter the
shape of the ET-VPD dependence we find the roots of the second
derivative of ET, using Eq. (\ref{m_n}) for stomatal conductance
($g_s$), with respect to VPD: \input{d2_solutions.tex} With this
result we have defined the family of curves separating concave up from
concave down ET solutions (Fig. \ref{concave}). These curves are
only functions of the exponent of the VPD dependence and a quantity we
call non-dimensional VPD ($VPD^m/g*$). Several important
relations reveal themselves from Eq. (\ref{curves}):

@@

While it is possible that in the future some other form of
VPD dependence is derived, at present \cite{MEDLYN_2011} and
\cite{Zhou_2014} firmly established n=m=1/2 as the most likely
candidate given current theory and empirical data. Additionally, we
argue that a concave up result matches physical intuition more than a
concave down result. Plants must maintain nutrient and sugar transport
through the phloem and xylem. To accomplish this, stomata must remain
slightly open \citep{De_2013, Nikinmaa_2013, Ryan_2014}. Furthermore,
even if complete stomatal closure were possible, cuticular water loss
and [at the ecosystem-scale] direct soil evaporation are still sources
of ET which increase with VPD, independent of stomatal
closure ** cite katul??? **. Therefore, in the limit as VPD becomes large and we assume
plants are exercising all strategies to reduce ET, any further increase
in VPD should result in an increase in ET through cuticular water loss
and/or direct soil evaporation. This inevitable transition from
conditions when stomata respond strongly to VPD to conditions when
stomata response is asymptoting towards full closure would cause a
concave up ET-VPD curve, which is matched by our theory. In short,
plant response becomes more limited as VPD increases, while
atmospheric demand monotonically increases with VPD, leading to the
 result that atmospheric demand dominates plant response when
atmospheric demand is high.

This analysis allows us to understand the theoretical shape of the ET
response to VPD with environmental conditions held
fixed. Accomplishing this with purely statistical methods applied to
flux observations would be very difficult, given the relatively fast
time scale of plant response and the non-stationarity of [solar
forced] environmental conditions over the relatively coarse (half
hourly) flux estimates (which is required to obtain robust
eddy-covariance statistics). Our results on the shape of the ET-VPD
curve with environmental conditions held fixed can be built upon with
future work examining how changes in VPD and environmental conditions
(e.g. soil water storage) feedback upon one another. In the soil water
storage example, over very long time scales extremely high VPD
perturbations coupled with no precipitation could result in decreases
in soil water storage such that water becomes limiting. This could be
represented by an extension of our framework in which uWUE is allowed
to change with changing SWC. Here, we focus our results by framing
theme with previous literature to build baseline intuition for ET-VPD
dependence.
